### 2022-10-02 manuscript/manuscript.tex
more writing

\documentclass[11pt]{article}
\RequirePackage{fullpage}
\RequirePackage[font=small,labelfont=bf]{caption}
\RequirePackage{amsmath,amssymb,amsthm}
\RequirePackage{graphicx}
\RequirePackage[hidelinks]{hyperref}
\RequirePackage{subcaption}
\RequirePackage{wasysym}
\RequirePackage{authblk}
\RequirePackage{bm}
\RequirePackage{bbm}
\RequirePackage[osf]{mathpazo}
\RequirePackage[bibstyle=authoryear,citestyle=authoryear-comp,
                date=year,
                maxbibnames=9,maxnames=5,maxcitenames=2,
                backend=biber,uniquelist=false,uniquename=false,
                % style=apa,
                sorting=nyt,
                % sorting=,
                hyperref=true]{biblatex}
\usepackage{color}
\usepackage{nicefrac}

@@

\section*{Theory}

@@

% First, we will show
% how the variance in allele frequency is related to the k

@@

\begin{align}
  \V(x_2) &= \E\left[(x_2 - x_0)^2\right] \\
          &= \E_1\left[\E_2\left[((x_2-x_1) + (x_1 - x_0))^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[(x_2-x_1)^2|x_1\right]\right] + \E_1\left[ (x_1 - x_0)^2\right] \\
          &= \frac{x_0(1-x_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{x_0(1-x_0)}{2N}
\end{align}

such that in general,

@@

Now, if the effective population size were to change each generation, equation
\eqref{eq:var_freq} would be,

\begin{align}
  \V(x_t) &= \V(x_0) \left(1 + \sum_{i=1}^t \prod_{j=1}^{i} \left(1 - \frac{1}{2N(j)}\right) \right)
\end{align}

Note that this too converges to $x_0(1-x_0)$ as $t \to \infty$.

Note now that equation 4 is a special case, when the evolutionary process has
independent allele frequency changes through the generations. Selection
processes will often lead to correlations in $\delta_i$ and $\delta_j$ for $i
\ne j$. These correlations are because the neutral allele in generations $i$
and $j$ can share some fraction of their fitness background, and in so far as
the selective pressures are the same in generations $i$ and $j$, selection will
perturb the frequency trajectories in the same direction.

One way we can model these dependencies is to partition $\delta_i$ into drift
and selection components. Over evolutionary replicates, the drift components
are independent and the selection components have some degree of
autocorrelation. Consider $\delta x_i = \delta_S x_i + \delta_D x_i$

\begin{align}
  \V(x_2) &= \V\left(\delta x_2 + \delta x_1\right) \\
          &= \V(\delta_D x_2) + \V(\delta_D x_1) + 2 \cov(\delta_S x_2, \delta_S x_1) \\
          &= \frac{x_0(1-x_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{x_0(1-x_0)}{2N} +  2 \cov(\delta_S x_2, \delta_S x_1) \\
\end{align}

since the drift changes are the familiar ones of equation XXX. The covariances
are determined by the selective process dynamics, linkage disequilibrium, and
the recombination rate.

Let us imagine there is some stochastic process $b(t)$ that rescales the coalescent
rate,

@@

\begin{align}
  \Delta_H p &= \langle x (f-1) \rangle \\
  \Delta_N p &= \langle x d \rangle \\
  \Delta_M p &= \langle \delta \rangle \\
\end{align}

@@

\begin{align}
  V(f) &\approx s^2 \V_a(n) + \delta_\text{LD}
\end{align}

>>>>

\documentclass[11pt]{article}
\RequirePackage{fullpage}
%\RequirePackage[font=small,labelfont=bf]{caption}
\RequirePackage{amsmath,amssymb,amsthm}
\RequirePackage{graphicx}
\RequirePackage[hidelinks]{hyperref}
\RequirePackage{subcaption}
\RequirePackage{wasysym}
\RequirePackage{authblk}
\RequirePackage{bm}
\RequirePackage{bbm}
%\RequirePackage[osf]{mathpazo}
\let\temp\rmdefault
\RequirePackage{mathpazo}
\let\rmdefault\temp

\RequirePackage[bibstyle=authoryear,citestyle=authoryear-comp,
                date=year,
                maxbibnames=9,maxnames=5,maxcitenames=2,
                backend=biber,uniquelist=false,uniquename=false,
                % style=apa,
                sorting=nyt,
                % sorting=,
                hyperref=true]{biblatex}
\usepackage{color}
\usepackage{nicefrac}

@@

\section*{Results}

\subsection*{Theory}

- overview of why BGS models fail?

% could be in intro too:

As genomes are transmitted through the generations, lifetime events selectively
alter what is passed to the next generation. While selection truly acts on the
genetic variants that directly cause heritable fitness differences among
individuals, neighboring regions experience an indirect contraction in genetic
variability. This byproduct is known as \emph{linked selection}, and it occurs
when selection changes a site's frequency on timescale faster than
recombination can decouple that site from its genetic background. Because
linked selection modifies the genetic variability in a regionally-correlated
way, it creates patterns in genomic data that are used to understand the types
of selection processes acting on genomes.

Much work has been done on developing mathematical theory for two primary
selection processes: hitchhiking and background selection. In turn, statistical
methods have use this theory for a variety of wide-ranging empirical problems.
Early on, researchers found correlations between recombination rates and
genetic diversity in small genomes with limited recombination like
\emph{Drosophila} (XXX), consistent with linked selection processes acting on
genome-wide scales. With population genomic data, small-scale (XXX kb) dips in
genetic diversity was evidence of rare but strong selective sweeps in humans
(XXX) and other species (XXX).

models of background selection can
closely predict levels of human-chimp divergence at the megabase scale (XXX).

humans (XXX), though these offered only a partial picture of the selection
processes acting (XXX).

Linked
selection studies have found additional strong selective sweeps in taxa across
the kingdom of life (XXX), and attention turned towards estimating the relative
contributions of the hitchhiking and background processes (Eyalshiv).
processes (Eyalshiv).

From an evolutionary perspective, models of linked selection

The other use of these models is to understand how linked selection processes
differentially alter the genetic variability, or equivalently, the genealogical
trees, along the genome.

Early on, some pointed out that evidence of strong selective sweeps in the
human genome was not the full story

Generally, in organisms with large amounts of recombination, linked selection
has

Weak selection processes

distorts
genome-wide

diversity.

The linked selection literature has primarily focused on developing theory for,
and quantifying the effects of two linked selection processes: hitchhiking and
background selection. Genome-wide models of the hitchhiking process consider
how beneficial mutations that sporadically enter the population

- QGLS polygenic linked selection

- H and W

- inflation factor

\subsection*{Statistical Methods}

\section*{Appendix}

\subsection*{Overview of Theory}

@@

\subsubsection*{Variance in Neutral Allele Frequency in a Neutral System}

@@

\begin{align}
  \V(x_2) &= \E\left[(x_2 - x_0)^2\right] \\
          &= \E_1\left[\E_2\left[((x_2-x_1) + (x_1 - x_0))^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[(\Delta x_2 + \Delta x_1)^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + 2\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right] + \E_1\left[ {\Delta x_1}^2\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + \E_1\left[ {\Delta x_1}^2\right] \\
          &= \frac{x_0(1-x_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{x_0(1-x_0)}{2N}
\end{align}

Note that the cross-moment $\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right]
= 0$ under neutral evolution. In general, the variance in allele frequency
after $t$ generations of neutral evolution is,

@@

% Now, if the effective population size were to change each generation, equation
% \eqref{eq:var_freq} would be,

% \begin{align}
%   \V(x_t) &= \V(x_0) \left(1 + \sum_{i=1}^t \prod_{j=1}^{i} \left(1 - \frac{1}{2N(j)}\right) \right)
% \end{align}

% Note that this too converges to $x_0(1-x_0)$ as $t \to \infty$.

\subsubsection*{Variance in Neutral Allele Frequency with Linked Selection}

Note that equation XXX is a special case, when the evolutionary process has
independent allele frequency changes through the generations and thus the
cross-moment is zero. Selection generates correlations between allele frequency
changes at nearby neutral sites. These correlations created by linked selection
inflate the variance in neutral allele frequency $\V(x_t)$ beyond what it would
be in a system without linked selection operating. The magnitude of this
inflation term is set by the selection process and its dynamics, linkage
disequilibrium, and the recombination rate. Even k

The theory of this began with \textcite{Robertson1961-ho},
and was thoroughly worked out by Santiago and Caballero
(\citeyear{Santiago1995-hx}, \citeyear{Santiago1998-bs}; see also
\cite{Barton2000-zg} and \cite{Buffalo2019-qs}). Our BGS statistical methods
employ the extension of this model for weak BGS \parencite{Santiago2016-mu}.
For completeness, we step through a quick derivation of the linked selection
inflation factor models here, based on
\textcite{Santiago1995-hx,Santiago1998-bs}.

Quantitative genetic models of linked selection

%%%%%

One way we can model these dependencies is to partition $\Delta_i$ into drift
and selection components. Over evolutionary replicates, the drift components
are independent and the selection components have some degree of
autocorrelation. Let us partition $\Delta x_i = \Delta_S x_i + \Delta_D x_i$,

\begin{align}
  \V(x_2) &= \V\left(\Delta x_2 + \Delta x_1\right) \\
          &= \V(\Delta_D x_2) + \V(\Delta_D x_1) + 2 \cov(\Delta_S x_2, \Delta_S x_1) \\
          &= \frac{x_0(1-x_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{x_0(1-x_0)}{2N} +  2 \cov(\Delta_S x_2, \Delta_S x_1) 
\end{align}

since $\Delta_D x_i \perp \Delta_D x_j$ for $i \ne j$. The covariances are
determined by the selective process dynamics, linkage disequilibrium, and the
recombination rate. These covariances are throughly worked out in
\textcite{Santiago1995-hx} and \textcite{Buffalo2019-qs}, so we give a short
derivation here. Throughout, we assume random mating, hermaphroditic
individuals, and a constant population size. We will use angular brackets to
indicate an average over all diploid individuals in the population, of some
value with an individual-level index. For example, if we denote the number of
allele an individual carries in generation $t$ as $z_{t,i} \in \{0, 1, 2\}$,
the population frequency at time $x_t$ is

\begin{align}
  x_t = \langle z_t \rangle = \frac{1}{2N} \sum_{i=1}^N z_{t,i}
\end{align}

Across evolutionary replicates, each parent $i$ passes one of their two neutral
alleles to a random number $k_i$ offspring. The \emph{heritable fitness} of an
individual $f_i$ is defined as the expected number of offspring passed down
across evolutionary replicates, such that $k_i = f_i + \varepsilon_i$.  Some
selection process determines the distribution of $f_i$, but the this occurs
independently of the process that sorts neutral alleles into diploid
individuals. We are unsure of the neutral allele count in an individual in the
initial generation, $z_{t,0}$, so we condition on it and take expectations
later. Then, the frequency in the initial and first generations are,

\begin{align}
  x_0 = \langle z_0 \rangle = \frac{1}{2N} \sum_{i=1}^N z_{0,i} \\
  x_1 = \langle z_1 \rangle = \frac{1}{2N} \sum_{i=1}^N \left( k_i \frac{z_{0,i}}{2} + \sum_{j=1}^{k_i} \delta_{i,j} \right)
\end{align}

The first term is the average number of neutral alleles each parent passes to
its $k_i$ offspring. In reality, alleles are discrete, so each heterozygous
parent also passes a random $\pm \nicefrac{1}{2}$ to each offspring, which
leads to the second term. The variable $\delta_{i,j}$ is 0 if the parent is a
homozygote, but if the parent is a heterozygote, is $\pm \nicefrac{1}{2}$ with
equal probability over evolutionary replicates.

Since individuals are exchangeable with their contribution to the
heterozygosity term and population size is constant,

\begin{align}
  \sum_{i=1}^N \sum_{j=1}^{k_i} \delta_{i,j} = \sum_{i=1}^N \delta_{i} 
\end{align}

We can further partition the offspring number into heritable and non-heritable
parts,

\begin{align}
  x_1 &= \frac{1}{2N} \sum_{i=1}^N \left( (f_i + \varepsilon_i) \frac{z_{0,i}}{2} + \delta_{i} \right) \\
      &= \frac{1}{4N} \sum_{i=1}^N f_i z_{0,i} + \frac{1}{4N} \sum_{i=1}^N \varepsilon_i z_{0,i} + \frac{1}{2N} \sum_{i=1}^N \delta_{i} \\
      &=  \frac{1}{4} \langle f \; z_{0} \rangle + \frac{1}{4} \langle \varepsilon \; z_{0} \rangle + \frac{1}{2}\langle \delta \rangle \\
\end{align}

% TODO factor 1/2 error

\begin{align}
  x_1 - x_0 &= \frac{1}{4} \langle f \; z_{0} \rangle + \frac{1}{4} \langle \varepsilon \; z_{0} \rangle + \frac{1}{2}\langle \delta \rangle - \langle z_0 \rangle \\
            &= \frac{1}{4} \langle (f-1) \; z_{0} \rangle + \frac{1}{4} \langle \varepsilon \; z_{0} \rangle + \frac{1}{2}\langle \delta \rangle 
\end{align}

Now, we look at the variance of this over evolutionary replicates. By symmetry,
the expected change in a neutral allele is $\E(\Delta x_t) = 0$ since we track
an arbitrary neutral allele. Then the variance $\V(\Delta x_1) = \E( {\Delta
x_1}^2)$,

\begin{align}
  \V(\Delta x_1) &= \frac{1}{4} \langle (f-1) \; z_{0} \rangle + \frac{1}{4} \langle \varepsilon \; z_{0} \rangle + \frac{1}{2}\langle \delta \rangle 
\end{align}

OLD
Let us imagine there is some stochastic process $b(t)$ that rescales the coalescent
rate,

@@

\begin{align}
  \Delta_H p &= \langle x (f-1) \rangle \\
  \Delta_N p &= \langle x d \rangle \\
  \Delta_M p &= \langle \Delta \rangle \\
\end{align}

@@

\begin{align}
  V(f) &\approx s^2 \V_a(n) + \Delta_\text{LD}
\end{align}
